2025 US School Shooting Predictions and Solutions

### 2024-11-28 paper/paper.qmd
update intro, discussion and result section

---
title: "2025 US School Shooting Predictions and Solutions"
subtitle: "Leveraging Forecasting to Shape Policy and Protect Communities"
author: 
  - Yun Chu
thanks: "Code and data are available at: https://github.com/chuyun2024/School-Shooting-Analysis."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

library(tidyverse)
library(ggplot2)
library(dplyr)
library(maps)
library(knitr)
library(randomForest)
library(here)
#### Read data ####
data <- read_csv("../data/02-analysis_data/analysis_data.csv")
```

# Introduction

Overview paragraph

Estimand paragraph

Results paragraph

Why it matters paragraph

Telegraphing paragraph: The remainder of this paper is structured as follows. @sec-data....

Paragraph of Estimand!

@@

## Overview

@@

- tidyverse [@tidyverse2024]
- dplyr [@dplyr2024]
- lubridate [@lubridate2024]
- readr [@readr2024]
- stringr [@stringr2024]
- arrow [@arrow2024]
- testthat [@testthat2024]
- randomForest [@randomForest2024]
- ggplot2
- maps
- knitr

## Measurement

@@

## Summary Statistics & Relationship Between Variables

@@

## Outcome variables

@@

## Predictor variables

@@

# Model

## Model Set-Up

@@

## Model Justification
- **Year** is included to account for temporal trends in school shootings, reflecting potential increases or decreases over time.
- **State** is treated as a categorical variable to capture location-specific effects, such as differences in legislation, policing, or socioeconomic factors.

@@

### Assumptions and Limitations
- **Assumptions**:
  1. Casualties are conditionally independent given the predictors.
  2. The relationship between predictors and casualties can be approximated by the ensemble of decision trees.
  3. Data is representative of underlying trends and free of major sampling biases.

@@

### Software Implementation
The model is implemented in **R**, using the randomForest [@randomforest] package for training and evaluation.

### Model Validation
- **Train/Test Splits**:
  The data is split into training (80%) and test (20%) sets. The model is trained on the former and validated on the latter to ensure generalizability.

@@

### Alternative Models Considered
- **Linear Regression**:
  - Strengths: Simplicity and interpretability.
  - Weaknesses: Inadequate for capturing non-linear relationships and interactions present in the data.

@@

# Results

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| label: fig-predicted_shooting_visualization
#| tbl-cap: "Predicted 2025 US School Shootings Numbers"
#| warning: false

@@

```{r}
#| echo: false
#| eval: false
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

modelsummary::modelsummary(
  list(
    "First model" = first_model
  ),
  statistic = "mad",
  fmt = 2
)
```

# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

Please don't use these as sub-heading labels - change them to be what your point actually is.

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

>>>>

---
title: "2025 US School Shooting Predictions and Solutions"
subtitle: "Leveraging Forecasting to Shape Policy and Protect Communities"
author: 
  - Yun Chu
thanks: "Code and data are available at: https://github.com/chuyun2024/School-Shooting-Analysis."
date: today
date-format: long
abstract: "School shootings in the United States pose a critical challenge with far-reaching impacts. This paper employs a Random Forest regression model to predict state-level casualties for 2025 using historical data from 1999 to the present. The findings reveal a positively skewed distribution of casualties, with states like Georgia, California, and Pennsylvania projected to face the highest numbers. These insights highlight the need for targeted interventions, including enhanced school safety measures and stricter gun control policies. By providing actionable predictions, this study aims to inform policies that reduce school shooting casualties and protect vulnerable populations."
format: pdf
toc: true
bibliography: references.bib
---

@@

library(tidyverse)
library(ggplot2)
library(dplyr)
library(maps)
library(knitr)
library(randomForest)
library(here)
library(kableExtra)
#### Read data ####
data <- read_csv("../data/02-analysis_data/analysis_data.csv")
```

# Introduction {#sec-intro}

School shootings remain a grave concern in the United States, with devastating consequences for students, educators, and communities. Despite the growing public discourse and policy debates surrounding school safety, little progress has been made in effectively predicting and preventing these tragic incidents. Understanding the patterns and predictors of school shootings is essential for designing evidence-based interventions to protect students and reduce casualties. This paper contributes to this critical issue by leveraging data-driven methodologies to analyze and predict school shooting casualties, with the aim of informing policy and prevention efforts.

The estimand in this paper is the number of casualties from school shootings across U.S. states for the year 2025. Using historical data from The Washington Post [@washingtonpost_school_shootings], which provides a comprehensive record of school shootings, a Random Forest regression model was developed to predict casualties based on state and temporal features. The analysis includes summary statistics, spatial visualizations, and predictive modeling, providing actionable insights for policymakers. This study fills a critical gap in understanding the state-level distribution of school shooting casualties and offers data-driven recommendations for targeted interventions.

A key finding of this study is the uneven distribution of casualties across states, with some states, such as Georgia, California, and Pennsylvania, predicted to experience significantly higher casualties in 2025. This highlights the importance of state-specific policies and resource allocation. The analysis also reveals that the distribution of casualties is positively skewed, with most incidents involving relatively few casualties, but a small number of high-casualty events disproportionately contributing to the overall impact. This information underscore the need for targeted strategies to mitigate the effects of high-casualty incidents.

The importance of this study lies in its ability to provide a data-driven foundation for policy and prevention efforts. By identifying high-risk states and understanding the characteristics of high-casualty incidents, this paper aims to guide policymakers in allocating resources effectively and designing interventions that address the underlying factors contributing to school shootings.

The structure of this paper is as follows: @sec-data discusses the data, including its sources, measurements, and key variables; @sec-model outlines the model used for prediction, including its justification and validation; @sec-result presents the results, including summary statistics, spatial visualizations, and predictions for 2025; and @sec-discussion concludes with a discussion of the findings, limitations, and recommendations for future research and policy.

@@

## Overview {#sec-data-overview}

@@

- tidyverse [@tidyverse2024]
- dplyr [@dplyr2024]
- lubridate [@lubridate2024]
- readr [@readr2024]
- stringr [@stringr2024]
- arrow [@arrow2024]
- testthat [@testthat2024]
- randomForest [@randomForest2024]
- ggplot2 [@ggplot2]
- maps [@maps]
- knitr [@knitr]
- here [@here]
- kableExtra [@kableExtra]

## Measurement {#sec-data-measurement}

@@

## Summary Statistics & Relationship Between Variables {#sec-data-sumsta}

@@

## Outcome variables {#sec-data-outvariable}

@@

## Predictor variables {#sec-predic-variable}

@@

# Model {#sec-model}

## Model Set-Up {#sec-model-setup}

@@

## Model Justification {#sec-model-justification}
- **Year** is included to account for temporal trends in school shootings, reflecting potential increases or decreases over time.
- **State** is treated as a categorical variable to capture location-specific effects, such as differences in legislation, policing, or socioeconomic factors.

Number of trees is selected to be 300 to balance accuracy and prevent overfitting.

@@

### Assumptions and Limitations {#sec-model-assumplim}
- **Assumptions**:
  1. Casualties are conditionally independent given the predictors.
  2. The relationship between predictors and casualties can be approximated by the ensemble of decision trees.
  3. Data is representative of underlying trends and free of major sampling biases.

@@

### Software Implementation {#sec-model-softimp}
The model is implemented in **R**, using the randomForest [@randomforest] package for training and evaluation.

### Model Validation {#sec-model-modelvalidation}
- **Train/Test Splits**:
  The data is split into training (80%) and test (20%) sets. The model is trained on the former and validated on the latter to ensure generalizability.

@@

### Alternative Models Considered {#sec-model-altmodel}
- **Linear Regression**:
  - Strengths: Simplicity and interpretability.
  - Weaknesses: Inadequate for capturing non-linear relationships and interactions present in the data.

@@

# Results {#sec-result}

Since random forest is non-parametric, there is not parameter estimates. ????

## Summary Statistics {#sec-result-summstat}

@tbl-summary_statistics_table shows the summary statistics for predicted 2025 school shooting casualties. On average, each school shooting incident results in approximately 1.59 casualties. At least half of the incidents involve only one casualty. The standard deviation of 1.91 indicates that while many incidents cluster around the mean of 1.59, there is considerable spread, with some incidents resulting in significantly higher casualties.

```{r}
#| echo: false
#| eval: true
#| label: tbl-summary_statistics_table
#| tbl-cap: "Summary Statistics for 2025 US School Shooting Casualties"
#| warning: false

# Load cleaned data
clean_data <- read.csv("../data/02-analysis_data/no_outlier_data.csv")

# Calculate summary statistics for the 'casualties' column
summary_stats <- clean_data %>%
  summarise(
    Mean = mean(casualties, na.rm = TRUE),
    Median = median(casualties, na.rm = TRUE),
    `Standard Deviation` = sd(casualties, na.rm = TRUE)
  )

# Display the summary statistics as a table
summary_stats %>%
  kable(
    format = "latex", 
    booktabs = TRUE, 
  ) %>%
  kable_styling(
    latex_options = c("striped"), 
    full_width = FALSE
  )

## Predicted Casualties by State {#sec-result-predicted}

Using a Random Forest model, predictions for total casualties in each state for the year 2025 were generated. The predicted casualties were visualized geographically using a choropleth map in @fig-predicted_shooting_visualization. States such as California and Texas are predicted to experience the highest casualties, reflecting historical trends. 
```{r}
#| echo: false
#| eval: true
#| label: fig-predicted_shooting_visualization
#| fig-cap: "Predicted 2025 US School Shootings Numbers"
#| warning: false

@@

The predicted casualties for each state are summarized in @tbl-predicted_shooting_table. The table provides a clear and concise overview, sorted in descending order of predicted casualties to emphasize the most affected states.

```{r}
#| echo: false
#| eval: true
#| label: tbl-predicted_shooting_table
#| tbl-cap: "Predicted School Shooting Casualties by State for 2025"
#| warning: false

# Prepare the predictions table without the year column
prediction_table <- prediction_data %>%
  select(-year) %>%  # Remove the 'year' column
  arrange(desc(predicted_casualties)) %>%
  rename(State = state, "Predicted Casualties" = predicted_casualties)

# Display the table using kable
prediction_table %>%
  kable(format = "latex", booktabs = TRUE) %>%
  kable_styling(latex_options = c("striped"))  # Removed "hold_position"
```

# Discussion {#sec-discussion}

## Actions to mitigate the impact of high-casualty incidents {#sec-discussion-first-point}

The distribution of school shooting casualties reveals significant insights into the nature of these incidents. The mean number of casualties is higher than the median, indicating a positively skewed distribution. This suggests that while most school shooting incidents result in relatively few casualties, there are a small number of events that result in significantly higher numbers of casualties. These high-casualty incidents have a disproportionate impact and should be a focal point for policymakers. Understanding the causes and contexts of such incidents, such as the presence of high-density schools, specific socio-economic factors, or insufficient security measures, could help in designing targeted interventions. For instance, schools in high-risk areas could benefit from increased security personnel, improved mental health resources for students, and stricter firearm regulations. Government must prioritize strategies aimed at mitigating the impact of these high-casualty incidents to protect vulnerable populations.

## Targeted Interventions for High-Risk States {#sec-discussion-sec-point}

The state-level predictions of casualties, as visualized in the map and table, provide actionable insights for policymakers at both the state and federal levels. States such as Georgia, California, and Pennsylvania are projected to experience higher numbers of casualties in 2025. This highlights the need for state-specific policies and resource allocation. For example, states with higher predicted casualties could implement tailored programs such as increasing mental health support in schools, conducting safety drills, and investing in school infrastructure to enhance safety measures. Additionally, identifying regions with consistently higher casualties can help governments allocate resources more effectively and conduct targeted investigations to address root causes.

## Policy Recommendations to Reduce Casualties {#sec-discussion-third-point}

The map and table of predicted casualties highlight opportunities for governments to take preventative action. One key recommendation is to introduce universal background checks and stricter gun control measures in states with high predicted casualties. Furthermore, funding programs that promote community engagement and early intervention in at-risk populations could help reduce the likelihood of school shootings. Schools can also be equipped with better surveillance technology and emergency response systems to minimize casualties during incidents. At a broader level, creating national frameworks for school safety policies can ensure consistency and accountability in protecting students and staff across all states.

## Weaknesses and next steps {#sec-discussion-weaknessnext}

While the Random Forest model provided reasonable predictions, it exhibited some limitations. The negative percentage of variance explained during initial iterations indicated that the model struggled to generalize with the available predictors. Although this was addressed by refining the data and removing outliers, the inclusion of only year and state as predictors limits the model's explanatory power. Additionally, the dataset's small size and imbalance across states might have impacted the model's ability to identify robust patterns.

The findings from this study underscore the importance of integrating predictive modeling with proactive policy-making. However, future work should focus on enhancing the dataset by incorporating additional predictors such as gun ownership rates, local crime statistics, and school-specific characteristics.

Additionally, examining the effectiveness of existing policies in high-risk areas could provide valuable insights into which interventions are most impactful.

Finally, collaborative efforts between federal agencies, state governments, and local communities will be essential to addressing the systemic issues underlying school shootings and reducing the number of casualties in future incidents.
